Attrition in Massive Open Online Courses

### 2015-01-12 paper_v2.Rnw
Updating paper after first review

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
\toappear{%$^\ast$ All authors contributed equally. \\
Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
{\emph{L@S'15}}, March 14--15, 2015, Vancouver, Canada. \\
Copyright \copyright~2015 ACM ISBN/14/04...\$15.00. \\
DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

@@

% Make sure hyperref comes last of your loaded packages, 
% to give it a fighting chance of not being over-written, 
% since its job is to redefine many LaTeX commands.
\usepackage[pdftex]{hyperref}
\hypersetup{
pdftitle={SIGCHI Conference Proceedings Format},
pdfauthor={LaTeX},
pdfkeywords={SIGCHI, proceedings, archival format},
bookmarksnumbered,
pdfstartview={FitH},
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black,
breaklinks=true,
}

@@

\title{Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors}

\numberofauthors{2}
\author{
  \alignauthor Omitted for blind review\\
  \affaddr{Institution}\\
  \affaddr{Address}\\
  \email{Email}\\
%   \alignauthor Ren\'{e} F. Kizilcec$^\ast$\\
%     \affaddr{Department of Communication}\\
%     \affaddr{Stanford University}\\
%     \email{kizilcec@stanford.edu}\\
%   \alignauthor Sherif Halawa$^\ast$\\
%     \affaddr{Department of Electrical Engineering}\\
%     \affaddr{Stanford University}\\
%     \email{halawa@stanford.edu}\\
}

@@

\begin{abstract}
Research on attrition has a long tradition in the field of education. Large-scale open learning environments prompted a shift away from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation. A systematic analysis of predictors of and reported reasons for attrition in 17 massive open online courses revealed gender and geographic inequalities in persistence and performance (Study 1). Time was a critical obstacle for many learners, but satisfaction was only weakly related to persistence. A follow-up study with online learners who disengaged from the course (Study 2; $N = 756$) found self-ascribed successful learners to report higher levels of goal striving, feelings of social belonging, and growth mindset. Iterative coding of open responses about challenges learners faced while taking the course reiterated the prevalence of having not enough time (reported by 84\%). A low capacity for volitional control was identified as a large contributor to this issue. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

\keywords{Online learning, persistence, attrition, dropout, disengagement, massive open online courses, MOOC, psychological factors}

@@

<<test1, echo=TRUE>>=
@

@@

Educational environments have become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}. Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer grading \cite{kulkarni2013peer,cambre2014talkabout}). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} resemble those on online media platforms. This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\label{tab:s1sum}
\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
\small
\center
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

Research and theorizing on attrition in education has a rich history. This review is intended to serve as a foundation to build on with the current research. We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work. The focus of this brief review is on how ways of thinking about attrition have developed over the last decades.

@@

The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting. The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course. An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, i.e., the extent to which the student engages in goal-direct behaviors in the face of distraction \cite{corno1993role}. Hence, motivation alone is necessary but not sufficient for persistence. Students may fail to sustain efforts in the absence of strong self-regulatory skills.

\begin{description}
  \item[H1] Successful MOOC learners exhibit higher levels of goal striving than unsuccessful ones.
\end{description}

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2007question}.

\begin{description}
  \item[RQ1] How do prior learner characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from a MOOC?
  \item[H2] Successful MOOC learners exhibit a greater sense of social belonging than unsuccessful ones.
\end{description}

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}. Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (i.e., instilling a growth mindset) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}. The results highlight the powerful influence of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.

\begin{description}
  \item[RQ2a] What proportion of learners who disengaged were satisfied with their achievements in the course?
  \item[RQ2b] How does learners' behavior in the MOOC vary by their level of satisfaction?
  \item[H3] Successful MOOC learners have a more fluid conception of intelligence (i.e., more of a growth than fixed mindset) than unsuccessful ones.
\end{description}

@@

Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration. However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner. Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students. Persistence is thought to be a function of background characteristics (e.g., demographics), academic and environmental variables (e.g., study habits, financial resources, work and family obligations), and academic and psychological outcomes (e.g., GPA, satisfaction). The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).

Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs. Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning. And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse. An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}. An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.

@@

A survey of over one hundred learners who dropped out of a MOOC showed that the majority indicated having too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}. A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}. A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that the likelihood of dropout was lower for those who i. actively participated in the first week of the course; ii. served as an authority figure in the community on the forum, and iii. did not engage in a particular subcommunity on the forum \cite{yang2013turn}. In addition, a number of machine learning approaches yielded promising results for early dropout prediction (e.g. \cite{taylor2014likely,halawa2014dropout}), but fell short of offering insights into why a given learner would leave the course.

\begin{description}
\item[RQ3a] What reasons do learners report for disengaging from a MOOC?
\item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
\end{description}

@@

The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with a combination of self-report and behavioral data from seventeen diverse MOOCs. Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments. This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf. Study 2).

@@

Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}). For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. 
% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (6-item index, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (see reasons in Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, we opted for binary-choice items to encourage learners to consider each reason in turn. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop," which should reduce acquiescence bias relative to yes/no or true/false scales.} Geographic location was determined by IP address.

>>>>

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
\toappear{
Permission to make digital or hard copies of all or part of this work for personal or classroom use is 	granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
{\emph{L@S'15}}, March 14--15, 2015, Vancouver, Canada. \\
Copyright \copyright~2015 ACM ISBN/14/04...\$15.00. \\
DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

@@

% Make sure hyperref comes last of your loaded packages, 
% to give it a fighting chance of not being over-written, 
% since its job is to redefine many LaTeX commands.
\usepackage[pdftex]{hyperref}
\hypersetup{
pdftitle={Attrition in Massive Open Online Courses},
pdfauthor={Rene F. Kizilcec, Sherif Halawa},
pdfkeywords={SIGCHI, proceedings, archival format},
bookmarksnumbered,
pdfstartview={FitH},
colorlinks,
citecolor=black,
filecolor=black,
linkcolor=black,
urlcolor=black,
breaklinks=true,
}

@@

\title{Attrition in Massive Open Online Courses}

\numberofauthors{2}
\author{
  \alignauthor Ren\'{e} F. Kizilcec\\
    \affaddr{Department of Communication}\\
    \affaddr{Stanford University}\\
    \email{kizilcec@stanford.edu}\\
  \alignauthor Sherif Halawa\\
    \affaddr{Department of Electrical Engineering}\\
    \affaddr{Stanford University}\\
    \email{halawa@stanford.edu}\\
}

@@

\begin{abstract}
Attrition in online learning is generally higher than in traditional settings. High attrition in large-scale open learning environments prompted a shift from a monolithic notion of success toward a more nuanced treatment that accounts for the diversity in learners' motivation.
A systematic analysis of individual differences in attrition in 17 diverse massive open online courses ($N>60,000$) revealed substantial achievement gaps: women were less likely to persist than men, and learners in Africa, Asia, and South America persisted less and scored lower than learners in Europe, North America, and Oceania. More persistent learners were, however, only marginally more satisfied with their achievement. 
Successful online learners who left the course early reported higher levels of goal striving, feelings of social belonging, and growth mindset than unsuccessful ones ($N=756$). The shortage of time was the primary obstacle for most learners, and many were hindered by insufficient volitional control.
A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their goals.
\end{abstract}

\keywords{Online learning, MOOC, persistence, dropout, individual differences, volition, social belonging, mindset}

@@

@@

Educational environments have become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost, with lower admission criteria, and with more flexible schedules to accommodate those who cannot afford to be full-time students \cite{goldrick2010challenges}. Distance learning programs deliver education to remote parts of the world and to people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (correspondence education), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

The latest generation of online learning environments, characterized by massive open online courses (MOOCs), has pushed the boundary on the scale of education \cite{waldrop2013campus}. By design, MOOCs provide course materials to millions of people worldwide. This scale could be achieved by pre-recording lectures, designing assessments that can be graded automatically, and by leveraging the momentum of the number of people involved (e.g., to facilitate peer learning or peer assessment). Maybe by virtue of their large scale, their prominent instructors, or their adherence to contemporary interface designs, MOOCs rapidly became an online media phenomenon. People would sign up weeks in advance of the course launch date, many of whom would never even enter the course site. And among those who enroll and enter the site, a large proportion tends to only ``sample'' some content and leave again \cite{kizilcec2013deconstructing}. Many of the prototypical behaviors observed in MOOCs \cite{kizilcec2013deconstructing,breslow2013studying} resemble those on online media platforms. This trend has also been reflected in the diversity of MOOC learners' motivations for enrolling \cite{kizilcec2015motivation}.

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. For instance, MOOCs can be thougth to have bridged two worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'' \cite{tyack1994grammar}, which presupposes instructor-defined goals that students strive to achieve \cite{kizilcec2015motivation}. This view of MOOC participation clearly adds complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses overall. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who reported having not enough time, and examined goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

\begin{table*}[th]
\caption{Summary statistics of learner demographics and behavior in 17 MOOCs}
\small
\center
\label{tab:s1sum}
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & 1,282    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & 630      & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & 2,116    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & 352      & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & 342      & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & 182      & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & 167      & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & 127      & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & 262      & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & 1,066    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & 40       & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & 510      & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 4,096   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & 282      & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 1,521   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & 610      & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & 626      & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (upper, lower quartile)  $^2$ bachelor's or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

Research and theorizing on attrition in education has a rich history. This review is intended to serve as a foundation to build on with the current research. We therefore simultaneously develop research questions and hypotheses to investigate and test in the current work. The focus of this brief review is on how ways of thinking about attrition have evolved over the last decades.

@@

The majority of early work on attrition centered around theoretical models of students' decision to persist or dropout of a traditional higher education setting. The literature on attrition in higher education is largely concerned with students disengaging from a course of study rather than a single course. An early model suggested that students' presistence is largely driven by their prior behavior, attitudes, and norms \cite{fishbein1975belief}. The psychological processes involved in turning an intent to learn into the decision to persist were thought to be mediated by volition, the extent to which the student engages in goal-directed behavior in the face of distraction \cite{corno1993role}. Motivation alone is necessary but not sufficient for persistence, because students may fail to sustain efforts in the absence of strong self-regulatory skills.

\begin{description}
  \item[H1] Successful learners exhibit higher levels of goal striving than unsuccessful learners.
\end{description}

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While prior experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments (e.g. \cite{walton2007question}).

\begin{description}
  \item[RQ1] How do prior characteristics (demographics, geographic location, intentions, and prior experience) influence learners' likelihood of disengaging from the course?
  \item[H2] Successful learners exhibit a greater sense of social belonging than unsuccessful learners.
\end{description}

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}. Nevertheless, a brief psychological intervention that taught community college students an incremental theory of intelligence (instilling a growth mindset \cite{dweck2006mindset}) was found to halve attrition rates and increase academic achievement \cite{paunesku2012brief}. This highlights the critical role of mindset and caution against overinterpreting non-persistence as an indicator of goal achievement.

\begin{description}
  \item[RQ2a] What proportion of learners who disengaged were satisfied with their achievements in the course?
  \item[RQ2b] How does learners' level of satisfaction depend on their behavior in the course?
  \item[H3] Successful learners have a more growth mindset than unsuccessful learners.
\end{description}

@@

Distance education, in contrast to traditional in-person education, attracts a different student demographic and typically provides fewer opportunities for social integration. However, students in distance learning programs tend to lead social lives outside of school, maybe working part-time and living with their partner. Building on Tinto's model, Bean and Metzner \citeyear{bean1985conceptual} proposed a conceptual model of persistence that would be more applicable to nontraditional students. Persistence is thought to be a function of background characteristics (demographics, etc.), academic and environmental variables (study habits, financial resources, work and family obligations, etc.), and academic and psychological outcomes (GPA, satisfaction, etc.). The significant change from Tinto's orginal model was the inclusion of environmental variables to account for the added complexity of nontraditional students' lives.
% (see \cite{kember1989longitudinal}, for another adaptation of Tinto's model for distance education).

Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs. Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning. And yet, these increasingly complex theoretical models were difficult to apply to real-world settings, and hence, the empirical evidence to support them remained sparse. An exploratory study of reasons for attrition of over 1,000 online students found eight relevant factors: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}. An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.

@@

A survey of over a hundred learners who dropped out of a MOOC showed that the majority indicated having too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}. A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}. A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that the likelihood of dropout was lower for those who i. actively participated in the first week of the course; ii. served as an authority figure in the community on the forum, and iii. did not engage in a particular subcommunity on the forum \cite{yang2013turn}. In addition, a number of machine learning approaches yielded promising results for early dropout prediction (e.g. \cite{taylor2014likely,halawa2014dropout}), but fell short of offering insights into why a given learner would leave the course.

\begin{description}
\item[RQ3a] What reasons do learners report for disengaging from a course?
\item[RQ3b] How do reasons for disengaging vary across prior learner characteristics (cf. RQ1)?
\end{description}

@@

The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with a combination of self-report and behavioral data from 17 diverse MOOCs. Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments. This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf. Study 2).

@@

Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (cf. Table \ref{tab:s1sum}). For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. 
% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (6-item index, range from 1 to 7, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (Figure \ref{fig:s1reason} lists reasons).\footnote{Instead of a select-all-that-apply question, learners were encouraged to consider each reason in turn by using binary-choice items. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to mitigate acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.
